---
title: "Random Forest"
author: "Cube Statistica"
date: "2022-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(rsample)
```

## Data Loading

```{r}
dataRaw <- read_csv("Data-DS-C1-Course.csv")
head(dataRaw, n = 10)

```
## Selecting Variables

We model Payment as our dependent variable and Age, Gender as our independent variable.
Before moving on, we select our required data.

```{r}
dataSelect <- select(dataRaw, Age, Gender, Payment)
head(dataSelect)
```

## Data Cleaning
Before Modelling for Random Forest, we will have to clean our data.
Firstly, we replace "Under 17" as 16 in Age feature.
Then we will convert Age as numeric feature and Gender as factor variable.
Then we replace all paid students as 1 and others as 0. This will make our
Payment column as Binary.

```{r}
# replace Under 17 with 16
dataSelect$Age <- str_replace(string = dataSelect$Age, pattern = "Under 17", replacement = "16")

# convert Age as numeric and Gender as factor
dataSelect$Age <- as.numeric(dataSelect$Age)
dataSelect$Gender <- as.factor(dataSelect$Gender)

# replace paid as 1 and others as non paid
dataSelect$Payment <- if_else(condition = dataSelect$Payment == "Paid", true = 1, false = 0)

# replace NA in payment as 0.
dataSelect$Payment <- dataSelect$Payment %>% replace_na(0)

tail(dataSelect)
```

```{r}
# Visualize the count of Gender Payments
ggplot(data = dataSelect, mapping = aes(x = Payment, fill = Gender)) +
  geom_bar()
```

```{r}
cross_table <- table(dataSelect$Payment, dataSelect$Gender)
cross_table <- kable(prop.table(cross_table),digits = 3) %>%
   kable_styling(latex_options = "striped",font_size = 20)
cross_table
```
## Split Data in Train and Test set

```{r}
# fix seed for randomness
set.seed(123)

# make outcome variable as factor since it is required by tidymodel
dataSelect$Payment <- as.factor(dataSelect$Payment)

# train-test split
datasplit <- initial_split(data = dataSelect, prop = 0.80)

# Get the resultant data set
data_train <- training(datasplit)
data_test <- testing(datasplit)
```

There are three steps for specifying the model using tidy verse :
1. Specify the type of model based on its mathematical structure (e.g., linear regression, random forest, KNN, etc).
2. Specify the engine for fitting the model.
3. When required, declare the mode of the model (mode for numeric outcome = regression)

```{r}
# Specify the model using parsnip in tidy model
rf_model <- rand_forest(trees = 1000, min_n = 5) %>% 
  set_engine("ranger", verbose = TRUE) %>% 
  set_mode("classification")

# Now we can use fit() or fit_xy()
rf_model_fit <- rf_model %>%
  fit(Payment ~ Age + Gender, data = data_train)
```

```{r}
# Make full test set prediction
model_prediction <- predict(object = rf_model_fit, new_data = data_test)

predicted.classes <- model_prediction$.pred_class
observed.classes <- data_test$Payment
```

```{r}
# accuracy
accuracy <- mean(observed.classes == predicted.classes)
accuracy
```
```{r}
# Confusion matrix, number of cases
table(observed.classes, predicted.classes)
```
```{r}
# Confusion matrix, proportion of cases
table(observed.classes, predicted.classes) %>% 
  prop.table() %>% round(digits = 3)
```




